---
title: "Figure 1 Vignette"
author: "Jianhong Ou"
vignette: >
  %\VignetteIndexEntry{Figure 1 Vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
output:
  html_document:
  theme: simplex
  toc: true
  toc_float: true
  toc_depth: 4
  fig_caption: true
---

```{r setup, include=FALSE}
suppressPackageStartupMessages({
  library(geomeTriD)
  library(geomeTriD.documentation)
  library(GenomicRanges)
  library(TxDb.Hsapiens.UCSC.hg19.knownGene)
  library(trackViewer)
})
knitr::opts_chunk$set(echo = TRUE)
```

## Figure 1

The Figure 1 is the showcase for `geomeTriD` package to present multiple genomic 
signals along with 3D models in different layout. 

### Load Libraries

```{r loadLibrary}
library(geomeTriD)
library(geomeTriD.documentation)
library(GenomicRanges)
library(TxDb.Hsapiens.UCSC.hg19.knownGene)
library(trackViewer)
```

### Prepare the annotation and genomic data
The data were downloaded from [ENCODE](https://www.encodeproject.org/) for GM12878 with assembly hg19.
The 3D structure model were downloaded from [FLAMINGO](https://github.com/wangjr03/FLAMINGO) results for GEO dataset with accession
[GSE63525](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE63525).

```{r readdata}
chr <- 'chr4'
range <- as(seqinfo(TxDb.Hsapiens.UCSC.hg19.knownGene)[chr], 'GRanges')
prefix <- 'https://www.encodeproject.org/files/'
url <- c(#ctcf = 'ENCFF364OXN/@@download/ENCFF364OXN.bigWig',
         #smc3 = 'ENCFF235BXX/@@download/ENCFF235BXX.bigWig',
         #rad21 = 'ENCFF567EGK/@@download/ENCFF567EGK.bigWig',
         #H3K4me3 = 'ENCFF674QZB/@@download/ENCFF674QZB.bigWig',
         #H3K27me3 = 'ENCFF594HSG/@@download/ENCFF594HSG.bigWig',
         H3K9me3 = 'ENCFF776OVW/@@download/ENCFF776OVW.bigWig',
         #H3K4me1 = 'ENCFF682WPF/@@download/ENCFF682WPF.bigWig',
         H3K27ac = 'ENCFF180LKW/@@download/ENCFF180LKW.bigWig')
urls <- paste0(prefix, url)
names(urls) <- names(url)
colorCode <- c('active'='#EA262E', 'inactive'='#179281')
# download the files and import the signals
genomicSigs <- importSignalFromUrl(urls, range = range,
                                   cols = list(
                                     'H3K9me3'=c('#000011', 'blue'),
                                     'H3K27ac'=c('#111100', 'orange')),
                                   format = 'BigWig')
# import the A/B compartments
compartments <- rtracklayer::import('https://ftp.ncbi.nlm.nih.gov/geo/series/GSE63nnn/GSE63525/suppl/GSE63525%5FGM12878%5Fsubcompartments.bed.gz')
compartments$col <- ifelse(is.na(compartments$name), 'gray', 
                           ifelse(grepl('^A', compartments$name),
                                  colorCode['active'],
                                  colorCode['inactive']))
compartments$label <- compartments$name
compartments$type <- 'compartment'
# import the 3D model generated by FLAMINGO
chrs <- paste0('chr', c(1:22, 'X'))
gm12878 <- paste0('https://github.com/wangjr03/FLAMINGO/',
                  'raw/refs/heads/main/predictions/GM12878/',
                  chrs, '_5kb.txt')
names(gm12878) <- chrs
gm12878.gl <- importFLAMINGO(gm12878)
```

### Plot the data by `geomeTriD`
```{r viewStructure}
# create a list of threeJsGeometry objects
objsC <- view3dStructure(gm12878.gl[[chr]], # 3D model by FLAMINGO
                         k=3, # 3D
                         feature.gr=compartments[seqnames(compartments) %in%
                                                   chr], # subset of the compartments
                         genomicSigs = genomicSigs, # the histon markers
                         reverseGenomicSigs = FALSE, lwd.maxGenomicSigs = 10,
                         show_coor = FALSE, label_gene = FALSE,#close the labels
                         renderer = 'none', # return a list of objects
                         resolution = 1)
## top vs bottom layout
for(i in seq_along(objsC)){
  if(grepl('backbone|compartment', names(objsC)[i])){
    objsC[[i]]$layer <- 'bottom' ## put compartment and backbone to bottom layer
  }
}
threeJsViewer(objsC, background = 'black', width = 1920, height = 1080)
# widget <- threeJsViewer(objsC, background = 'black', width = 1920, height = 1080)
# tempfile <- 'Fig1.part1.html'
# htmlwidgets::saveWidget(widget, file=tempfile)
# utils::browseURL(tempfile)

## side by side layout
for(i in seq_along(objsC)){
  if(grepl('backbone|compartment', names(objsC)[i])){
    objsC[[i]]$layer <- 'top' ## put compartment to top layer
  }
  if(grepl('H3K9me3|H3K27ac', names(objsC)[i])){
    objsC[[i]]$side <- 'right' ## put H3K9me3 and H3K27ac signals to right panel
  }
}
threeJsViewer(objsC,
              title = c('A/B compartment',
                        'H3K9me3 and H3K27Ac signals'),
              background = 'black', width = 1920, height = 1080)
# widget <- threeJsViewer(objsC,
#               title = c('A/B compartment',
#                         'H3K9me3 and H3K27Ac signals'),
#               background = 'black', width = 1920, height = 1080)
# tempfile <- 'Fig1.part2.html'
# htmlwidgets::saveWidget(widget, file=tempfile)
# utils::browseURL(tempfile)
```

## SessionInfo

```{r sessionInfo}
sessionInfo()
```