---
title: "Figure 4A Vignette"
author: "Jianhong Ou"
vignette: >
  %\VignetteIndexEntry{Figure 4A Vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
output:
  html_document:
  theme: simplex
  toc: true
  toc_float: true
  toc_depth: 4
  fig_caption: true
---

```{r setup, include=FALSE}
suppressPackageStartupMessages({
  library(geomeTriD)
  library(geomeTriD.documentation)
  library(GenomicRanges)
  library(TxDb.Hsapiens.UCSC.hg38.knownGene)
  library(org.Hs.eg.db)
  library(grid)
  library(colorRamps)
})
knitr::opts_chunk$set(echo = TRUE, out.width = "900px")
```

# Figure 4A

Figure 4A highlights the geomeTriD package, illustrating its ability to visualize 3D models derived from [Dip-C](https://doi.org/10.1126/science.aat5641), along with multiple genomic signals mapped onto single-cell 3D structures.

## Load Libraries

```{r loadLibrary}
library(geomeTriD)
library(geomeTriD.documentation)
library(GenomicRanges)
library(TxDb.Hsapiens.UCSC.hg38.knownGene)
library(org.Hs.eg.db)
library(colorRamps)
```

## Present single cell 3D structure for human Dip-C data

This data were downloaded from GEO with accession [GSE117874](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE117874).
We will first present the GM12878 cell 3 with all chromosomes.
```{r, out.width="500px"}
## set supperloop positions
supperloops <- GRanges(c('Xa:56800000-56850000',
                         'Xa:75350000-75400000',
                         'Xa:115000000-115050000',
                         'Xa:130850000-130900000',
                         'Xb:56800000-56850000',
                         'Xb:75350000-75400000',
                         'Xb:115000000-115050000',
                         'Xb:130850000-130900000'))
names(supperloops) <- c('ICCE', 'x75', 'DXZ4', 'FIRRE',
                        'ICCE', 'x75', 'DXZ4', 'FIRRE')
supperloops$label <- names(supperloops)
supperloops$col <- c(2:5, 2:5) ## set colors for each element
supperloops$type <- 'gene' ## set it as gene

## set the data folder, all data are available in the extdata folder of this package
extdata <- system.file('extdata', 'GSE117874', package='geomeTriD.documentation')
hickit_3dg <- dir(extdata, '3dg', full.names = TRUE)
## load the data
hickit <- import3dg(hickit_3dg, parental_postfix=NULL)[[1]]
## set colors to emphersize the Xa and Xb
col.backbone <- gray.colors(n = length(seqlevels(hickit)))
names(col.backbone) <- seqlevels(hickit)
col.backbone['Xa'] <- 'brown'
col.backbone['Xb'] <- 'orange'

cell3 <- view3dStructure(hickit, feature.gr = supperloops,
                renderer = 'none', resolution=1,
                col.backbone = col.backbone, lwd.backbone = 0.25)
backbone <- extractBackbonePositions(cell3, names(cell3)[grepl('backbone', names(cell3))])
supperloop_spheres <- createTADGeometries(supperloops, backbone, alpha = 0.8)
threeJsViewer(cell3, supperloop_spheres)
```


We will then present the four supperloop anchors in chrX of GM12878 cell with segment and sphere style.
```{r}
hickit <- import3dg(hickit_3dg, parental_postfix=c('a', 'b'))[[1]]
## split it into maternal and paternal 3D structures.
hickit.a <- hickit[hickit$parental=='a'] ## mat
hickit.b <- hickit[hickit$parental=='b'] ## pat
## delete the parental information
hickit.a$parental <- hickit.b$parental <- NULL
## set data range
range <- GRanges('X:1-155270560')
## prepare the coordinates for the four superloops. 
supperloops <- GRanges(c('X:56800000-56850000',
                      'X:75350000-75400000',
                      'X:115000000-115050000',
                      'X:130850000-130900000'))
names(supperloops) <- c('ICCE', 'x75', 'DXZ4', 'FIRRE')
supperloops$label <- names(supperloops)
supperloops$col <- 2:5 ## set colors for each element
supperloops$type <- 'gene' ## set it as gene
## subset the data by given range
hickit.a <- subsetByOverlaps(hickit.a, range)
hickit.b <- subsetByOverlaps(hickit.b, range)
hickit.a <- alignCoor(hickit.a, hickit.b)
## prepare the backbone colors
resolution <- 3
backbone_colors <- matlab.like2(n=resolution*length(hickit.a))
## add the superloops as segments
c1 <- view3dStructure(hickit.a,
                      feature.gr=supperloops,
                      renderer = 'none',
                      region = range,
                      resolution=resolution,
                      show_coor=FALSE,
                      lwd.backbone = 0.25,
                      col.backbone = backbone_colors,
                      lwd.gene=6)
c2 <- view3dStructure(hickit.b,
                      feature.gr=supperloops,
                      renderer = 'none',
                      region = range,
                      resolution=resolution,
                      show_coor=FALSE,
                      lwd.backbone = 0.25,
                      col.backbone = backbone_colors,
                      lwd.gene=6)
c2 <- lapply(c2, function(.ele) { ## put pat to right pannel
  .ele$side = 'right'
  .ele
})
## view the data
threeJsViewer(c1, c2, title = c('GM12878 cell 3 mat', 'GM12878 cell 3 pat'))
#widget <-threeJsViewer(c1, c2, title = c('mat', 'pat'))
#tempfile <- 'Fig4.html'
#htmlwidgets::saveWidget(widget, file=tempfile)
#utils::browseURL(tempfile)
```
```{r}
## view the superloops as spheres
addFeaturesAsSphere <- function(obj, features, ...){
  backbone <- extractBackbonePositions(obj)
  spheres <- createTADGeometries(features, backbone, ...)
  c(obj, spheres)
}
mat <- view3dStructure(hickit.a, renderer = 'none', region = range,
                       resolution=resolution, show_coor=FALSE,
                       lwd.backbone = 0.25,
                       col.backbone = backbone_colors)
mat <- addFeaturesAsSphere(mat, supperloops, alpha = 0.5)
pat <- view3dStructure(hickit.b, renderer = 'none', region = range,
                       resolution=resolution, show_coor=FALSE,
                       lwd.backbone = 0.25,
                       col.backbone = backbone_colors)
mat <- addFeaturesAsSphere(mat, supperloops, alpha = 0.5)
pat <- addFeaturesAsSphere(pat, supperloops, alpha = 0.5)
showPairs(mat, pat, title = c('GM12878 cell 3 mat', 'GM12878 cell 3 pat'))
```

### Plot all cells to show the cell heterogeneity
```{r}
## load the processed data.
hickit.a <- readRDS(file.path(extdata, 'hickit.a.rds'))
hickit.b <- readRDS(file.path(extdata, 'hickit.b.rds'))
widgets <- mapply(function(a, b, i){
  mat <- view3dStructure(a, renderer = 'none', region = range,
                         resolution=resolution, show_coor=FALSE, lwd.backbone = 0.25,
                      col.backbone = backbone_colors)
  mat <- addFeaturesAsSphere(mat, supperloops, alpha = 0.5)
  pat <- view3dStructure(b, renderer = 'none', region = range,
                         resolution=resolution, show_coor=FALSE, lwd.backbone = 0.25,
                      col.backbone = backbone_colors)
  pat <- addFeaturesAsSphere(pat, supperloops, alpha = 0.5)
  showPairs(mat, pat, title = paste('cell', i, c('mat', 'pat')),
            background = c("#11111188",
                           "#222222DD",
                           "#222222DD",
                           "#11111188"))
}, hickit.a, hickit.b, names(hickit.a), SIMPLIFY = FALSE)

widgets[[1]]
widgets[[2]]
```




## SessionInfo

```{r sessionInfo}
sessionInfo()
```