---
title: "Figure S5 Vignette"
author: "Jianhong Ou"
vignette: >
  %\VignetteIndexEntry{Figure S5 Vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
output:
  html_document:
  theme: simplex
  toc: true
  toc_float: true
  toc_depth: 4
  fig_caption: true
---

```{r setup, include=FALSE}
suppressPackageStartupMessages({
  library(geomeTriD)
  library(geomeTriD.documentation)
  library(GenomicRanges)
  library(grid)
  library(colorRamps)
  library(TxDb.Mmusculus.UCSC.mm10.knownGene)
  library(org.Mm.eg.db)
  library(geometry)
  library(GenomeInfoDb)
})
knitr::opts_chunk$set(echo = TRUE, out.width = "900px")
options(rgl.useNULL = TRUE)
```

# Figure S5

Figure S5 showcases the `geomeTriD` package, demonstrating how it presents 3D models along with multiple genomic signals mapped onto single-cell 3D structures.

## Load Libraries

```{r}
library(geomeTriD)
library(geomeTriD.documentation)
library(GenomicRanges)
library(GenomeInfoDb)
library(colorRamps)
library(TxDb.Mmusculus.UCSC.mm10.knownGene)
library(org.Mm.eg.db)
library(geometry)
```

## Present single cell 3D structure for ChAIR data

Here we present 3D genome models with integrated genomic signals derived from [Tri-omic](https://doi.org/10.1038/s41592-025-02658-7), also known as ChAIR (capture of chromatin Accessibility, Interaction, and RNA profiles). This method enables simultaneous mapping of chromatin accessibility, chromatin interactions, and gene expression. Paired-end tag (PET) data are used to reconstruct 3D genome structures, while the accompanying RNA-seq and ATAC-seq signals are visualized along these models.

```{r}
## plot region
range <- GRanges('chrX:1-20000000')
## prepare features
feature.gr <- getFeatureGR(txdb = TxDb.Mmusculus.UCSC.mm10.knownGene,
                           org = org.Mm.eg.db,
                           range = range)

## load the 3D structure for PET data
stage <- c('G1', 'S', 'G2M')
extdata <- system.file('extdata', 'ChAIR', package = 'geomeTriD.documentation')
ChAIR_PET_m <- import3dg(file.path(extdata, 'PET', paste0('patski.allele.', stage, '_2_m.pairs.chrX.3dg.gz')))
ChAIR_PET_p <- import3dg(file.path(extdata, 'PET', paste0('patski.allele.', stage, '_2_p.pairs.chrX.3dg.gz')))
## get the chrX coordinates for maternal and paternal
ChAIR_PET_m <- lapply(ChAIR_PET_m, function(.ele) {
  seqlevelsStyle(.ele) <- 'UCSC'
  .ele$parental <- NULL
  subsetByOverlaps(.ele, range)
  })
ChAIR_PET_p <- lapply(ChAIR_PET_p, function(.ele) {
  seqlevelsStyle(.ele) <- 'UCSC'
  .ele$parental <- NULL
  subsetByOverlaps(.ele, range)
})
# import the signals for ATAC-seq and RNA-seq for maternal and paternal
M <- lapply(stage, function(cc){ ## maternal
  importGenomicSigs(c(ATAC=file.path(extdata, 'ATAC', paste0('patski.allele.', cc, '_2.M.bw')),
                      RNA=file.path(extdata, 'RNA', paste0('patski.allele.', cc, '_2.M.bw'))),
                      range = range,
                    cols = c('darkgreen', 'darkred'),
                    format = 'BigWig')
  })
P <- lapply(stage, function(cc){ ## paternal
  importGenomicSigs(c(ATAC=file.path(extdata, 'ATAC', paste0('patski.allele.', cc, '_2.P.bw')),
                      RNA=file.path(extdata, 'RNA', paste0('patski.allele.', cc, '_2.P.bw'))),
                      range = range,
                    cols = c('darkgreen', 'darkred'),
                    format = 'BigWig')
})
## set backbone color, make it easy to catch the genomic coordinates
resolution <- 3
backbone_colors <- rev(rainbow(n=resolution*length(ChAIR_PET_m[[1]])))
## only label the genes with high expression
labelHighExpGene <- function(sig){
  this.feature.gr <- feature.gr
  expressed <- sig$RNA$dat[sig$RNA$dat$score>100]
  this.feature.gr$label[countOverlaps(feature.gr, expressed)==0] <- NA
  ## remove the genes with big size
  this.feature.gr$label[width(this.feature.gr)>100000] <- NA
  ## only show the first gene label if multiple high expressed gene
  ## in the same region
  l.rle <- rle(!is.na(this.feature.gr$label))
  l.rle$values[l.rle$lengths<=5] <- TRUE ## merge the region with small gap
  l.rle <- rle(inverse.rle(l.rle))
  l <- cumsum(l.rle$lengths)
  this.feature.gr$label[-l[l.rle$values]] <- NA
  return(this.feature.gr)
}

## create threeJsGeometry objects for maternal and paternal cells
M_cells <- mapply(ChAIR_PET_m, M, FUN=function(model, sig){
  cell <- view3dStructure(model,
                  feature.gr = labelHighExpGene(sig),
                  genomicSigs = sig,
                  signalTransformFun = log1p,
                  lwd.maxGenomicSigs = 8,
                  reverseGenomicSigs = FALSE,
                  renderer = 'none',
                  resolution = resolution,
                  col.backbone = rev(rainbow(n=resolution*length(model))),
                  lwd.backbone = 0.25,
                  lwd.gene = 1,
                  region = range,
                  show_coor = FALSE,
                  cluster3Dpoints = TRUE)
}, SIMPLIFY = FALSE)
P_cells <- mapply(ChAIR_PET_p, P, FUN=function(model, sig){
  cell <- view3dStructure(model,
                  feature.gr = labelHighExpGene(sig),
                  genomicSigs = sig,
                  signalTransformFun = log1p,
                  lwd.maxGenomicSigs = 8,
                  reverseGenomicSigs = FALSE,
                  renderer = 'none',
                  resolution = resolution,
                  col.backbone = rev(rainbow(n=resolution*length(model))),
                  lwd.backbone = 0.25,
                  lwd.gene = 1,
                  region = range,
                  show_coor = FALSE,
                  cluster3Dpoints = TRUE)
}, SIMPLIFY = FALSE)

## show the maternal and paternal structures side by side
showPairs(M_cells[[1]], P_cells[[1]], title=c('mat G1', 'pat G1')) #G1
showPairs(M_cells[[2]], P_cells[[2]], title=c('mat S', 'pat S')) #S
showPairs(M_cells[[3]], P_cells[[3]], title=c('mat G2M', 'pat G2M')) #G2M

## check the size 
getV <- function(points){
  vol <- convhulln(as.matrix(mcols(points)), options='Fa')$vol
}
sapply(ChAIR_PET_m, getV)
sapply(ChAIR_PET_p, getV)
```

## SessionInfo

```{r sessionInfo}
sessionInfo()
```